\chapter{Håstad's Switching Lemma}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overview}

This chapter formalises \textbf{Håstad's Switching Lemma} (1987) via Razborov's
decision-tree path encoding. Informally:

\begin{quote}
If $f$ is a width-$w$ DNF and $\rho$ is a uniformly random $s$-restriction,
then the probability that $f\!\restriction\!\rho$ requires a decision tree of
depth $>d$ is at most $(5sw/n)^d$ (or a constant-factor variant thereof).
\end{quote}

The Lean statement, \lean{SwitchingLemma2.switching_lemma}, is a combinatorial
counting inequality rather than a probability bound; the two are equivalent
after rearranging.  References: Håstad (1987); Razborov's simplified proof;
O'Donnell, \emph{Analysis of Boolean Functions}, Ch.~4.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The formalised statement}

\begin{theorem}[Switching Lemma -- Lean statement]
\label{thm:switching_lemma}
\lean{SwitchingLemma2.switching_lemma}
\uses{def:dnf_width,def:is_restriction,def:is_bad_restriction,
      def:num_s_restrictions,hyp:hnd,hyp:five_s_le_n}
Let $n>0$, let $f:\{0,1\}^n\to\{0,1\}$ be a DNF formula, and fix $w,s,d\in\bbn$.
Assume
\begin{itemize}
  \item $\text{width}(f)\le w$ (width counted as list length, see
        Section~\ref{sec:faithfulness}),
  \item $5s \le n$,
  \item $f$ satisfies the \emph{non-contradictory clause} hypothesis
        \texttt{hnd}: for every term $t\in f$ and every pair of literals
        $\ell_1,\ell_2\in t$, if $\ell_1.\mathrm{var}=\ell_2.\mathrm{var}$ then
        $\ell_1=\ell_2$.
\end{itemize}
Then
\[
  \bigl|\{\,\rho : \rho\text{ is an $s$-restriction and }
        \mathrm{dtDepth}(f|_\rho)>d\,\}\bigr| \cdot n^d
  \;\le\;
  \mathrm{numSRestrictions}(n,s)\,\cdot\,(10\,s\,w)^d.
\]
\end{theorem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Faithfulness critique}\label{sec:faithfulness}

The Lean statement is faithful to the classical lemma \emph{modulo} a handful
of subtle modelling choices.  This section documents each one so a user of
\lean{SwitchingLemma2.switching_lemma} knows precisely what they are signing
up for.

%-----------------------------------------------------------------------------
\subsection{How DNFs are modelled}

Recall (\lean{Literal}, \lean{Term}, \lean{DNF} in
\texttt{TCSlib/BooleanAnalysis/Circuit.lean}):
\begin{itemize}
  \item A \textbf{literal} is a pair $(v,\mathrm{neg})\in\Fin n\times\mathrm{Bool}$.
  \item A \textbf{term} is a \emph{list} of literals, not a set or a
        \texttt{Finset}.
  \item A \textbf{DNF} is a list of terms.
  \item \lean{Term.width} is defined as \texttt{List.length}, i.e.\ literal
        occurrences are \emph{counted with multiplicity}.
  \item \lean{DNF.width} is the (list-)maximum of the term widths.
\end{itemize}

\paragraph{Consequence 1 -- Width vs.\ literal count.}
In the textbook, ``width $w$'' means each term involves at most $w$
\emph{distinct} variables.  In our formalisation, \texttt{f.width ≤ w} is the
stronger statement ``each term has at most $w$ literal \emph{occurrences}''.
For a syntactically clean DNF (no repeated literals inside a clause) the two
notions coincide and the hypothesis is trivial to supply.  For a DNF with
repeated literals, the formalised width can be strictly larger, which means
the user must either dedup before applying the theorem or accept a looser
bound.  This is not an unsoundness, but it is a papercut that is worth
flagging.

\paragraph{Consequence 2 -- The \texttt{hnd} hypothesis.}
\label{hyp:hnd}
The parameter
\[
  \texttt{hnd} \;:\; \forall t\in f,\;\forall \ell_1,\ell_2\in t,\;
  \ell_1.\mathrm{var}=\ell_2.\mathrm{var}\;\Rightarrow\;\ell_1=\ell_2
\]
is \emph{not} the same as ``each term uses distinct variables''.  It only
forbids a term from containing both a literal and its negation (a
self-falsifying term).  It does \emph{allow} a term to contain the same
literal twice.  Two remarks:

\begin{enumerate}
  \item The classical statement does not explicitly require \texttt{hnd};
        self-falsifying clauses are implicitly removed as identically false.
        Our formalisation makes this requirement explicit, so the user must
        discharge a proof obligation that is mechanical but nonzero.  For a
        concrete DNF built from literals read off a file or a syntactic
        representation, \texttt{decide} usually suffices.
  \item \texttt{hnd} is used crucially in
        \lean{SwitchingLemma2.encode_go_not_kills_first_clause} and its
        callers to guarantee that, when the encoder processes the free-var
        literals of a term, the variable-to-literal lookup is well-defined
        (no variable appears with two different polarities in the same term).
        Without it, the round-trip property fails.
\end{enumerate}

%-----------------------------------------------------------------------------
\subsection{The restriction regime: $5s \le n$}
\label{hyp:five_s_le_n}

Classical Hastad is typically stated for any $s$, with the bound reading
$(5pw)^d$ where $p = s/n$.  The formalisation bakes in the hypothesis
$5s\le n$, i.e.\ $p\le 1/5$.  This is \emph{not} an artefact of the
combinatorics per se but of the particular inductive proof of the Vandermonde
identity (\lean{SwitchingLemma2.choose_mul_pow_bound}) used to convert
$\binom{n}{s-d}(4n)^d\le \binom{n}{s}(5s)^d$.  For restrictions with $p>1/5$
one must either generalise that lemma or apply the theorem vacuously via the
$d>s$ branch (\lean{bad_filter_empty_of_d_ge_s}).

\paragraph{$0<n$.}
Required so that $\Fin n$ is inhabited.  Trivial.

%-----------------------------------------------------------------------------
\subsection{The $(10sw)^d$ constant vs.\ $(5sw)^d$}

Razborov's encoding yields $(5sw/n)^d$ in the sharpest accounting.  The
formalised constant is $10sw$ rather than $5sw$.  The extra factor of two
comes from the combinatorial bound on the encoder image size
(\lean{aux_image_card_bound} via \lean{exists_aux_injection}), which tags
each of the $d$ encoded triples as $\Fin w\times\mathrm{Bool}\times\mathrm{Bool}$
(position, polarity, end-of-block marker).  A sharper encoder is possible but
was sidestepped for proof-engineering simplicity.  This is a loss of a
constant factor, not an asymptotic one.

%-----------------------------------------------------------------------------
\subsection{Status}

\begin{itemize}
  \item \lean{SwitchingLemma2.switching_lemma} is stated and the outer
        structure is complete.
  \item \lean{SwitchingLemma2.bad_count_bound} currently contains a
        \texttt{sorry} (line 636 of \texttt{MainTheorem.lean}) which is the
        summation step that passes from ``fibre bound'' to ``total bound''.
        Everything else is \texttt{sorry}-free.
  \item \lean{SwitchingLemma2.RoundTrip.go\_roundtrip\_gen} contains one
        inductive-step \texttt{sorry} at the re-entry point of the induction
        hypothesis.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Dependency graph}

The proof is layered across seven Lean files.  Information flows downward:

\begin{center}
\begin{tabular}{l|l}
\textbf{File} & \textbf{Key contents} \\ \hline
\texttt{Circuit.lean}
  & \lean{Literal}, \lean{Term}, \lean{DNF}, \lean{CNF},
    \lean{DecisionTree}, \lean{dtDepth}, \lean{buildFullDTree}.\\
\texttt{Switching/Defs.lean}
  & \lean{Restriction}, \lean{Restriction.extend}, \lean{numFree},
    \lean{freeVars}, \lean{Literal.killedBy}, \lean{Term.fixedBy},
    \lean{restrictFn}, \lean{IsBadRestriction}, \lean{IsRestriction},
    \lean{numSRestrictions}, \lean{first_clause_preserved}. \\
\texttt{Switching/CanonicalDTree.lean}
  & \lean{canonicalDTree} (Razborov's DT), correctness,
    \lean{dtDepth_restrictFn_le_numFree} (depth $\le$ numFree).\\
\texttt{Switching/Encoding.lean}
  & \lean{razborovEncode}, \lean{razborovDecode},
    \lean{processClauseLits}. \\
\texttt{Switching/EncodingProperties.lean}
  & Numerous invariants of the encoder; uses \texttt{hnd} for
    \lean{encode_go_not_kills_first_clause}. \\
\texttt{Switching/RoundTrip.lean}
  & \lean{go_roundtrip_gen} (decode $\circ$ encode = id),
    \lean{razborovEncode_injective}. \\
\texttt{Switching/MainTheorem.lean}
  & Encoder image parser, \lean{exists_aux_injection},
    \lean{aux_image_card_bound}, \lean{fiber_bound},
    \lean{bad_count_bound},
    \lean{choose_mul_pow_bound}, \lean{switching_lemma},
    \lean{switching_corollary}. \\
\end{tabular}
\end{center}

\noindent
Conceptually the proof factors as:
\[
\underbrace{
  \texttt{razborovEncode}\text{ is injective on bad restrictions}
}_{\text{RoundTrip.lean (needs hnd)}}
\;\Rightarrow\;
\underbrace{
  |\{\text{bad}\}| \le |\mathrm{image}(\texttt{razborovEncode})|
}_{\text{set-theoretic}}
\;\Rightarrow\;
\underbrace{
  |\mathrm{image}| \le (\text{combinatorial count})
}_{\text{MainTheorem.lean}}.
\]
The combinatorial count is obtained by fibring the image over the first
component $\gamma$ (a restriction with $s-d$ fewer free variables), bounding
each fibre by $(4w)^d$ (the aux-data count
\lean{aux_image_card_bound}), then summing fibres via the Vandermonde-type
inequality \lean{choose_mul_pow_bound}, which is where $5s\le n$ is consumed.

%-----------------------------------------------------------------------------
\section{Corollary: no small CNF approximations}

\begin{theorem}[Switching corollary]
\label{thm:switching_corollary}
\lean{SwitchingLemma2.switching_corollary}
\uses{thm:switching_lemma}
Under the same hypotheses as Theorem~\ref{thm:switching_lemma},
\[
  \bigl|\{\rho : \mathrm{IsRestriction}\ s\ \rho,\ f|_\rho
    \text{ has no CNF of width}\le w\}\bigr| \cdot n^w
  \le \mathrm{numSRestrictions}(n,s)\,(10sw)^w.
\]
\end{theorem}

\noindent
This uses the fact (\lean{dtDepth_le_implies_small_dnf_cnf}) that a function
with small decision-tree depth admits both a small-width DNF and a
small-width CNF.
